---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data, including dependent packages
```{r}

library(ggplot2);
library(plyr);
#Read activity file
activity <- read.csv("./activity.csv");activity <- read.csv("./activity.csv");
#Activity_withoutNA has no NA records
activity_wtna <- na.omit(activity,cols=("steps"));
```  
  
## What is mean total number of steps taken per day?  
  
```{r}
dsteps <- ddply(activity_wtna, .(date), summarize, steps=sum(steps));
hist(dsteps$steps,main="Histogram of total steps per day",xlab="Steps per day");
mean <- format(mean(dsteps$steps),digit=7);
median <- median(dsteps$steps);
```

  
The mean when ignoring NA records is `r mean`  
The median when ignoring NA records  is `r median`
  
## What is the average daily activity pattern?
```{r}
#Summarize data
isteps <- ddply(activity_wtna, .(interval), summarize, msteps=mean(steps,na.rm=TRUE));
#Plot time series
plot(isteps$interval,isteps$msteps,type="l",xlab="Interval",ylab="Steps",main="Steps by interval");
#Interval with the max avg steps across all days
minterval <- isteps[isteps$msteps == max(isteps$msteps,na.rm=TRUE),];
```

The maximum interval is `r minterval'.

## Imputing missing values

The code below use data frame merge to populate a msteps colum.
The msteps (mean of steps for interval i) column populates the steps column if
its value is NA.  
  
```{r}
# Count of NA values
nrow(activity[is.na(activity$steps),]);

#Fill NA by merging with isteps (average steps by interval)
activity_backfill <- merge(activity,isteps);

activity_backfill[is.na(activity_backfill$step),]$steps <- activity_backfill[is.na(activity_backfill$step),]$msteps; 

activity_backfill$msteps <- NULL;

#Create histogram of daily steps
dsteps_backfill <- ddply(activity_backfill, .(date), summarize, steps=sum(steps));
hist(dsteps_backfill$steps,main="Histogram of total steps per day",xlab="Steps per day");

mean_backfill <- format(mean(dsteps_backfill$steps),digit=7);
median_backfill <- format(median(dsteps_backfill$steps),digit=7);

```
      

The mean when ignoring NA records is `r mean`  
The median when ignoring NA records  is `r median`  

The mean with backfill of NA data is `r mean_backfill`  
The median with backfill of NA data  is `r median_backfill`  
  
After backfill, the mean stays the same and the median increases slightly.
  
## Are there differences in activity patterns between weekdays and weekends?

The code below use the qplot function of the ggplot2 package.

I use the $day property of the POSIXlt object to test if the day is a Sunday (0) or a Saturday (6) and generate a factor column added to the activity dataframe.

```{r}


# Create a boolean vector - TRUE means this is a week end
Wend <-  (as.POSIXlt(activity_backfill$date)$wday == 0 ) |  
  (as.POSIXlt(activity_backfill$date)$wday == 6 );

Wend <- as.data.frame(Wend);

#This function turn the boolen into either weekend or weekday
Wdayend <- function(x) {
  if (x == TRUE) {return ("weekend")}
  else 
  { return("weekday")}
}

#A column is added & converted into a factor
activity_backfill$Wend <- apply(Wend,1,Wdayend);
activity_backfill$Wend <- as.factor(activity_backfill$Wend);

#Summarize data by interval and Wend column (factor with "weekend" and "weekday" values)
isteps <- ddply(activity_backfill, .(interval,Wend), summarize, msteps=mean(steps,na.rm=TRUE));

#Plot time series
qplot(interval,msteps,data=isteps,facets= Wend~.,
      geom=c("line"),
      main="Comparison of mean steps by interval for Weekend and weekdays",
      ylab="Avg number of steps by interval");

```


